---
title: "emmeans: Estimated marginal means"
author: "Patrick Cherry"
date: "2023-05-02"
output:
  github_document: 
    toc: yes
    df_print: kable
  pdf_document: 
    toc: yes
    latex_engine: xelatex
    df_print: kable
  html_document:
    df_print: kable
    code_folding: show
    number_sections: no
    toc: yes
    toc_float: yes
mainfont: Helvetica Neue
---
```{r, setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)  # set default chunk option to print the code that generated that output
options(knitr.table.format = 'markdown') # render kable tables in md for github, comment out otherwise
library(tidyverse)
library(broom)
library(emmeans)
library(patchwork)                  # for combining plots
```
```{r}
vignette("models", package = "emmeans")
```

Estimated marginal means (EMMs, previously known as least-squares means in the context of traditional regression models) are derived by using a model to make predictions over a regular grid of predictor combinations (called a reference grid).

# The pigs dataset
## Initial model fitting
Consider the pigs dataset provided with the package (help("pigs") provides details). These data come from an experiment where pigs are given different percentages of protein (percent) from different sources (source) in their diet, and later we measured the concentration (conc) of leucine. The percent values are quantitative, but we chose those particular values deliberately, and (at least initially) we want separate estimates at each percent level; that is, we want to view percent as a factor, not a quantitative predictor.

As discussed, our first task is to come up with a good model. Doing so requires a lot of skill, and we don’t want to labor too much over the details; you really need other references to deal with this aspect adequately. But we will briefly discuss five models and settle on one of them:
```{r, warning = FALSE}
mod1 <- lm(conc ~ source * factor(percent), data = pigs)
par(mfrow = c(2,2)); plot(mod1)
```
```{r}
mod2 <- update(mod1, . ~ source + factor(percent))   # no interaction
par(mfrow = c(2,2)); plot(mod2)
```

```{r}
map_dfr(list(mod1, mod2), glance)
```


These models have R2 values of 0.808 and 0.700, and adjusted R2 values of 0.684 and 0.634. mod1 is preferable to mod2, suggesting we need the interaction term. However, a residual-vs-predicted plot of mod2 has a classic “horn” shape (curving and fanning out), indicating a situation where a response transformation might help better than including the interaction.

It turns out that an inverse transformation, (1/conc) really serves us well. (Perhaps this isn’t too surprising, as concentrations are typically determined by titration, in which the actual measurements are volumes; and these are reciprocally related to concentrations, i.e., amounts per unit volume.)

So here are three more models:
```{r}
mod3 <- update(mod1, inverse(conc) ~ .)
mod4 <- update(mod2, inverse(conc) ~ .)     # no interaction
mod5 <- update(mod4, . ~ source + percent)  # linear term for percent
par(mfrow = c(2,2)); plot(mod5)
```

```{r}
map_dfr(list(mod1, mod2, mod3, mod4, mod5), glance)
```
(Note: We could have used 1/conc as the response variable, but emmeans provides an equivalent inverse() function that will prove more advantageous later.) The residual plots for these models look a lot more like a random scatter of points (and that is good). The R2 values for these models are 0.818, 0.787, and 0.749, respectively; and the adjusted R2s are 0.700, 0.740, and 0.719. mod4 has the best adjusted R2 and will be our choice.

# Estimated marginal means
```{r}
(EMM.source <- emmeans(mod4, "source"))
```

```{r}
(EMM.percent <- emmeans(mod4, "percent"))
```

Calling `tidy()` (from `broom`) on the object will put it into a beautiful data frame. we could make a plot.

```{r}
tidy(EMM.percent)
```


## Comparison to ordinary means
Let’s compare these with the ordinary marginal means (OMMs) on inverse(conc):

```{r}
with(pigs, tapply(inverse(conc), source, mean))
```

Can I write the above ordinary means in Tidyverse/dplyr language? 
```{r}
pigs %>%
  mutate(conc = 1/conc) %>%
  summarize(mean = mean(conc), .by = source)
```

```{r}
with(pigs, tapply(inverse(conc), percent, mean))
```

```{r}
pigs %>%
  mutate(conc = 1/conc) %>%
  summarize(mean = mean(conc), .by = percent)
```

Both sets of OMMs are vaguely similar to the corresponding EMMs. However, please note that the EMMs for percent form a decreasing sequence, while the the OMMs decrease but then increase at the end.

# The reference grid, and definition of EMMs

Estimated marginal means are defined as marginal means of model predictions over the grid comprising all factor combinations – called the reference grid. For the example at hand, the reference grid is

```{r}
(RG <- expand.grid(source = levels(pigs$source), percent = unique(pigs$percent)))
```
